---
title: Manipulating, analyzing and exporting data with tidyverse
author: Data Carpentry contributors
minutes: 45
---

```{r, echo=FALSE, purl=FALSE, message=FALSE}
source("setup.R")
surveys <- read.csv("data/portal_data_joined.csv")
suppressWarnings(surveys$date <- lubridate::ymd(paste(surveys$year,
                                                      surveys$month,
                                                      surveys$day,
                                                      sep = "-")))
```

:::::::::::::::::::::::::::::::::::::::  objectives

- Describe the purpose of the **`dplyr`** and **`tidyr`** packages.
- Select certain columns in a data frame with the **`dplyr`** function `select`.
- Extract certain rows in a data frame according to logical (boolean) conditions with the **`dplyr`** function `filter` .
- Link the output of one **`dplyr`** function to the input of another function with the 'pipe' operator `%>%`.
- Add new columns to a data frame that are functions of existing columns with `mutate`.
- Use the split-apply-combine concept for data analysis.
- Use `summarize`, `group_by`, and `count` to split a data frame into groups of observations, apply summary statistics for each group, and then combine the results.
- Describe the concept of a wide and a long table format and for which purpose those formats are useful.
- Describe what key-value pairs are.
- Export a data frame to a .csv file.

:::::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::: questions

- What are dplyr and tidyr?
- How can I select specific rows and/or columns from a dataframe?
- How can I combine multiple commands into a single command?
- How can I create new columns or remove existing columns from a dataframe?

:::::::::::::::::::::::::::::::::::::::::::::::::::
# Loading packages 
Packages in R are basically sets of additional functions that let you do
more stuff. The functions we've been using so far, like `round()`,
`sqrt()`, or `c()` come built into R. Packages give you access to
additional functions beyond base R. A similar function to `read.csv()`
is `read_csv()` from the tidyverse package which we will mention in the next lesson. We don't have time to cover their differences but notice that the exact spelling
determines which function is used. Before you use a package for the
first time you need to install it on your machine, and then you should
import it in every subsequent R session when you need it.

Packages in our pb&j example is kind of like giving somebody a
box full of 'sandwich making' tools and machines.
Maybe you wanna give them a box pb&j machine, a bread maker, a bread knife
etc. All these things are related to making a sandwich, some of them,
like the bread knife is actually used in our theoretical pb&j machine.

Thats all packages R a related set of functions written and built by some
developer

To install the **`tidyverse`** package, we can type
`install.packages("tidyverse")` straight into the console. In fact, it's
better to write this in the console than in our script for any package,
as there's no need to re-install packages every time we run the script.
Then, to load the package type:

```{r, message=FALSE, purl=FALSE}
## load the tidyverse packages, incl. dplyr
library(tidyverse)
```

Now we can use the functions from the **`tidyverse`** package. Let's use
`read.csv()` to read the data into a data frame (we will learn more
about data frames later):
# Data manipulation using **`dplyr`** and **`tidyr`**

Bracket subsetting is handy, but it can be cumbersome and difficult to read,
especially for complicated operations. Enter **`dplyr`**. **`dplyr`** is a package for
helping with tabular data manipulation. It pairs nicely with **`tidyr`** which enables you to swiftly convert between different data formats for plotting and analysis.

The **`tidyverse`** package is an
"umbrella-package" that installs **`tidyr`**, **`dplyr`**, and several other useful packages for data analysis, such as  **`ggplot2`**, **`tibble`**, etc.

The **`tidyverse`** package tries to address 3 common issues that arise when
doing data analysis in R:

1. The results from a base R function sometimes depend on the type of data.
2. R expressions are used in a non standard way, which can be confusing for new
  learners.
3. The existence of hidden arguments having default operations that new learners are not aware
  of.

You should already have installed and loaded the **`tidyverse`** package.
If you haven't already done so, you can type `install.packages("tidyverse")` straight into the console. Then, type `library(tidyverse)` to load the package.

## What are **`dplyr`** and **`tidyr`**?

The package **`dplyr`** provides helper tools for the most common data manipulation
tasks. It is built to work directly with data frames, with many common tasks
optimized by being written in a compiled language (C++). An additional feature is the
ability to work directly with data stored in an external database. The benefits of
doing this are that the data can be managed natively in a relational database,
queries can be conducted on that database, and only the results of the query are
returned.

This addresses a common problem with R in that all operations are conducted
in-memory and thus the amount of data you can work with is limited by available
memory. The database connections essentially remove that limitation in that you
can connect to a database of many hundreds of GB, conduct queries on it directly, and pull
back into R only what you need for analysis.

The package **`tidyr`** addresses the common problem of wanting to reshape your data for
plotting and usage by different R functions. For example, sometimes we want data sets where we have one
row per measurement. Other times we want a data frame where each measurement type has its
own column, and rows are instead more aggregated groups
(e.g., a time period, an experimental unit like a plot or a batch number).
Moving back and forth between these formats is non-trivial, and **`tidyr`** gives you tools
for this and more sophisticated  data manipulation.

To learn more about **`dplyr`** and **`tidyr`** after the workshop, you may want to check out this
[handy data transformation with **`dplyr`** cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-transformation.pdf)
and this [one about **`tidyr`**](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-import.pdf).

As before, we'll read in our data using the `read_csv()` function from the
tidyverse package **`readr`**.

```{r, results="hide", purl=FALSE}
surveys <- read_csv("data/portal_data_joined.csv")
```

```{r, results="hide", purl=FALSE}
## inspect the data
str(surveys)
```

```{r, eval=FALSE, purl=FALSE}
## preview the data
view(surveys)
```

Next, we're going to learn some of the most common **`dplyr`** functions:

- `select()`: subset columns
- `filter()`: subset rows on conditions
- `mutate()`: create new columns by using information from other columns
- `group_by()` and `summarize()`: create summary statistics on grouped data
- `arrange()`: sort results
- `count()`: count discrete values

## Selecting columns and filtering rows
<!-- Go through these slowly and make sure people understand -->
To select columns of a data frame, use `select()`. The first argument
to this function is the data frame (`surveys`), and the subsequent
arguments are the columns to keep.

```{r, results="hide", purl=FALSE}
select(surveys, plot_id, species_id, weight)
```

To select all columns *except* certain ones, put a "-" in front of
the variable to exclude it.

```{r, results="hide", purl=FALSE}
select(surveys, -record_id, -species_id)
```

This will select all the variables in `surveys` except `record_id`
and `species_id`.

To choose rows based on a specific criterion, use `filter()`:

```{r, purl=FALSE}
filter(surveys, year == 1995)
```

## Pipes

What if you want to select and filter at the same time? There are three
ways to do this: use intermediate steps, nested functions, or pipes.

With intermediate steps, you create a temporary data frame and use
that as input to the next function, like this:

```{r, purl=FALSE}
surveys2 <- filter(surveys, weight < 5)
surveys_sml <- select(surveys2, species_id, sex, weight)
```

This is readable, but can clutter up your workspace with lots of objects that you have to name individually. With multiple steps, that can be hard to keep track of.

You can also nest functions (i.e. one function inside of another), like this:

```{r, purl=FALSE}
surveys_sml <- select(filter(surveys, weight < 5), species_id, sex, weight)
```

This is handy, but can be difficult to read if too many functions are nested, as
R evaluates the expression from the inside out (in this case, filtering, then selecting).

The last option, *pipes*, are a recent addition to R. Pipes let you take
the output of one function and send it directly to the next, which is useful
when you need to do many things to the same dataset.  Pipes in R look like
`%>%` and are made available via the **`magrittr`** package, installed automatically
with **`dplyr`**. If you use RStudio, you can type the pipe with <kbd>Ctrl</kbd>

- <kbd>Shift</kbd> + <kbd>M</kbd> if you have a PC or <kbd>Cmd</kbd> +
  <kbd>Shift</kbd> + <kbd>M</kbd> if you have a Mac.

```{r, purl=FALSE}
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
```

In the above code, we use the pipe to send the `surveys` dataset first through
`filter()` to keep rows where `weight` is less than 5, then through `select()`
to keep only the `species_id`, `sex`, and `weight` columns. Since `%>%` takes
the object on its left and passes it as the first argument to the function on
its right, we don't need to explicitly include the data frame as an argument
to the `filter()` and `select()` functions any more.

Some may find it helpful to read the pipe like the word "then." For instance,
in the example above, we took the data frame `surveys`, *then* we `filter`ed
for rows with `weight < 5`, *then* we `select`ed columns `species_id`, `sex`,
and `weight`. The **`dplyr`** functions by themselves are somewhat simple,
but by combining them into linear workflows with the pipe we can accomplish
more complex manipulations of data frames.

If we want to create a new object with this smaller version of the data, we
can assign it a new name:

```{r, purl=FALSE}
surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

surveys_sml
```

Note that the final data frame is the leftmost part of this expression.

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge {.challenge}

Using pipes, subset the `surveys` data to include animals collected before
1995 and retain only the columns `year`, `sex`, and `weight`.

:::::::: solution

```{r, answer=TRUE, eval=FALSE, purl=FALSE}
surveys %>%
    filter(year < 1995) %>%
    select(year, sex, weight)
```

:::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, eval=FALSE, purl=TRUE, echo=FALSE}
## Pipes Challenge:
##  Using pipes, subset the data to include animals collected
##  before 1995, and retain the columns `year`, `sex`, and `weight.`
```

### Mutate

Frequently you'll want to create new columns based on the values in existing
columns, for example to do unit conversions, or to find the ratio of values in two
columns. For this we'll use `mutate()`.

To create a new column of weight in kg:

```{r, purl=FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000)
```

You can also create a second new column based on the first new column within the same call of `mutate()`:

```{r, purl=FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000,
         weight_lb = weight_kg * 2.2)
```

If this runs off your screen and you just want to see the first few rows, you
can use a pipe to view the `head()` of the data. (Pipes work with non-**`dplyr`**
functions, too, as long as the **`dplyr`** or `magrittr` package is loaded).

```{r, purl=FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
```

The first few rows of the output are full of `NA`s, so if we wanted to remove
those we could insert a `filter()` in the chain:

```{r, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
```

`is.na()` is a function that determines whether something is an `NA`. The `!`
symbol negates the result, so we're asking for every row where weight *is not* an `NA`.

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge {.challenge}

Create a new data frame from the `surveys` data that meets the following
criteria: contains only the `species_id` column and a new column called
`hindfoot_cm` containing the `hindfoot_length` values (currently in mm)
converted to centimeters.
In this `hindfoot_cm` column, there are no `NA`s and all values are less
than 3.

**Hint**: think about how the commands should be ordered to produce this data frame!

:::::::: solution

```{r, answer=TRUE, eval=FALSE, purl=FALSE}
surveys_hindfoot_cm <- surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    mutate(hindfoot_cm = hindfoot_length / 10) %>%
    filter(hindfoot_cm < 3) %>%
    select(species_id, hindfoot_cm)
```

:::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, eval=FALSE, purl=TRUE, echo=FALSE}
## Mutate Challenge:
##  Create a new data frame from the `surveys` data that meets the following
##  criteria: contains only the `species_id` column and a new column called
##  `hindfoot_cm` containing the `hindfoot_length` values converted to centimeters.
##  In this `hindfoot_cm` column, there are no `NA`s and all values are less
##  than 3.

##  Hint: think about how the commands should be ordered to produce this data frame!
```

### Split-apply-combine data analysis and the `summarize()` function

Many data analysis tasks can be approached using the *split-apply-combine*
paradigm: split the data into groups, apply some analysis to each group, and
then combine the results. Key functions of **`dplyr`** for this workflow are
`group_by()` and `summarize()`.

#### The `group_by()` and `summarize()` functions

`group_by()` is often used together with `summarize()`, which collapses each
group into a single-row summary of that group.  `group_by()` takes as arguments
the column names that contain the **categorical** variables for which you want
to calculate the summary statistics. So to compute the mean `weight` by sex:

```{r, purl=FALSE}
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```

You may also have noticed that the output from these calls doesn't run off the
screen anymore. It's one of the advantages of `tbl_df` over data frame.

You can also group by multiple columns:

```{r, purl=FALSE}
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>%
  tail()
```

Here, we used `tail()` to look at the last six rows of our summary. Before, we had
used `head()` to look at the first six rows. We can see that the `sex` column contains
`NA` values because some animals had escaped before their sex and body weights
could be determined. The resulting `mean_weight` column does not contain `NA` but
`NaN` (which refers to "Not a Number") because `mean()` was called on a vector of
`NA` values while at the same time setting `na.rm = TRUE`. To avoid this, we can
remove the missing values for weight before we attempt to calculate the summary
statistics on weight. Because the missing values are removed first, we can omit
`na.rm = TRUE` when computing the mean:

```{r, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight))
```

Here, again, the output from these calls doesn't run off the screen
anymore. If you want to display more data, you can use the `print()` function
at the end of your chain with the argument `n` specifying the number of rows to
display:

```{r, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight)) %>%
  print(n = 15)
```

Once the data are grouped, you can also summarize multiple variables at the same
time (and not necessarily on the same variable). For instance, we could add a
column indicating the minimum weight for each species for each sex:

```{r, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight))
```

It is sometimes useful to rearrange the result of a query to inspect the values. For instance, we can sort on `min_weight` to put the lighter species first:

```{r, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(min_weight)
```

To sort in descending order, we need to add the `desc()` function. If we want to sort the results by decreasing order of mean weight:

```{r, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(desc(mean_weight))
```

#### Counting

When working with data, we often want to know the number of observations found
for each factor or combination of factors. For this task, **`dplyr`** provides
`count()`. For example, if we wanted to count the number of rows of data for
each sex, we would do:

```{r, purl=FALSE}
surveys %>%
    count(sex)
```

The `count()` function is shorthand for something we've already seen: grouping by a variable, and summarizing it by counting the number of observations in that group. In other words, `surveys %>% count()` is equivalent to:

```{r, purl=FALSE}
surveys %>%
    group_by(sex) %>%
    summarize(count = n())
```

For convenience, `count()` provides the `sort` argument:

```{r, purl=FALSE}
surveys %>%
    count(sex, sort = TRUE)
```

Previous example shows the use of `count()` to count the number of rows/observations
for *one* factor (i.e., `sex`).
If we wanted to count *combination of factors*, such as `sex` and `species`,
we would specify the first and the second factor as the arguments of `count()`:

```{r, purl=FALSE}
surveys %>%
  count(sex, species)
```

With the above code, we can proceed with `arrange()` to sort the table
according to a number of criteria so that we have a better comparison.
For instance, we might want to arrange the table above in (i) an alphabetical order of
the levels of the species and (ii) in descending order of the count:

```{r, purl=FALSE}
surveys %>%
  count(sex, species) %>%
  arrange(species, desc(n))
```

From the table above, we may learn that, for instance, there are 75 observations of
the *albigula* species that are not specified for its sex (i.e. `NA`).

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge {.challenge}

1. How many animals were caught in each `plot_type` surveyed?

:::::::: solution

```{r, answer=TRUE, purl=FALSE}
surveys %>%
    count(plot_type)
```

:::::::::::::::::

2. Use `group_by()` and `summarize()` to find the mean, min, and max hindfoot
  length for each species (using `species_id`). Also add the number of
  observations (hint: see `?n`).

:::::::: solution

```{r, answer=TRUE, purl=FALSE}
surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    group_by(species_id) %>%
    summarize(
        mean_hindfoot_length = mean(hindfoot_length),
        min_hindfoot_length = min(hindfoot_length),
        max_hindfoot_length = max(hindfoot_length),
        n = n()
    )
```

:::::::::::::::::

3. What was the heaviest animal measured in each year? Return the columns `year`,
  `genus`, `species_id`, and `weight`.

:::::::: solution

```{r, answer=TRUE, purl=FALSE}
surveys %>%
    filter(!is.na(weight)) %>%
    group_by(year) %>%
    filter(weight == max(weight)) %>%
    select(year, genus, species, weight) %>%
    arrange(year)
```

:::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, eval=FALSE, purl=TRUE, echo=FALSE}
## Count Challenges:
##  1. How many animals were caught in each `plot_type` surveyed?

##  2. Use `group_by()` and `summarize()` to find the mean, min, and max
## hindfoot length for each species (using `species_id`). Also add the number of
## observations (hint: see `?n`).

##  3. What was the heaviest animal measured in each year? Return the
##  columns `year`, `genus`, `species_id`, and `weight`.
```


# Exporting data

Now that you have learned how to use **`dplyr`** to extract information from
or summarize your raw data, you may want to export these new data sets to share
them with your collaborators or for archival.

Similar to the `read_csv()` function used for reading CSV files into R, there is
a `write_csv()` function that generates CSV files from data frames.

For this workshop we're going to save things in the same folder: `data`,
but remember it's good practice to keep them separate. Ideally you
should create a `data_raw` folder that should only contain the raw,
unaltered data. So that even if the files it contains are deleted, we can always
re-generate them.

In preparation for our next lesson on plotting, we are going to prepare a
cleaned up version of the data set that doesn't include any missing data.

Let's start by removing observations of animals for which `weight` and `hindfoot_length` are missing, or the `sex` has not been determined:

```{r, purl=FALSE}
surveys_complete <- surveys %>%
  filter(!is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         !is.na(sex))                # remove missing sex
```

Because we are interested in plotting how species abundances have changed
through time, we are also going to remove observations for rare species (i.e.,
that have been observed less than 50 times). We will do this in two steps: first
we are going to create a data set that counts how often each species has been
observed, and filter out the rare species; then, we will extract only the
observations for these more common species:

```{r, purl=FALSE}
## Extract the most common species_id
species_counts <- surveys_complete %>%
    count(species_id) %>%
    filter(n >= 50)

## Only keep the most common species
surveys_complete <- surveys_complete %>%
  filter(species_id %in% species_counts$species_id)
```

```{r, eval=FALSE, purl=TRUE, echo=FALSE}
### Create the dataset for exporting:
##  Start by removing observations for which the `species_id`, `weight`,
##  `hindfoot_length`, or `sex` data are missing:
surveys_complete <- surveys %>%
    filter(species_id != "",        # remove missing species_id
           !is.na(weight),                 # remove missing weight
           !is.na(hindfoot_length),        # remove missing hindfoot_length
           sex != "")                      # remove missing sex

##  Now remove rare species in two steps. First, make a list of species which
##  appear at least 50 times in our dataset:
species_counts <- surveys_complete %>%
    count(species_id) %>%
    filter(n >= 50) %>%
    select(species_id)

##  Second, keep only those species:
surveys_complete <- surveys_complete %>%
    filter(species_id %in% species_counts$species_id)
```

To make sure that everyone has the same data set, check that `surveys_complete`
has `r nrow(surveys_complete)` rows and `r ncol(surveys_complete)` columns by
typing `dim(surveys_complete)`.

Now that our data set is ready, we can save it as a CSV file in our `data`
folder.

```{r, purl=FALSE, eval=FALSE}
write_csv(surveys_complete, file = "data/surveys_complete.csv")
```

```{r, purl=FALSE, eval=TRUE, echo=FALSE}
if (!dir.exists("data")) dir.create("data")
write_csv(surveys_complete, file = "data/surveys_complete.csv")
```


::::::::::::::::::::::::::::::::::::: keypoints

- Use the `dplyr` package to manipulate data frames.
- Use `select()` to choose variables from a data frame.
- Use `filter()` to choose data based on values.
- Use `mutate()` to create new variables.
- Use `group_by()` and `summarize()` to work with subsets of data.

::::::::::::::::::::::::::::::::::::::::::::::::
